Network Origins of Social Distinction

### 2025-07-20 manuscript.tex

The first aspect of social distinction is \textit{status}. Status refers to the internal stratification of all network members, along with the connections of deference and obligation. In the case of politics, status refers to the ability to access those who themselves have higher levels of access, with those most central to the access network having the authority to make decisions. This can be captured traditionally by measures of status such as eigenvector centrality \citep{bonacich1972norms}. Status is higher for actors when the people they are connected to are themselves influential, and accessibility is merely a political reframe of the context. The internal stratification of individuals is determined by who has greater connections to those with even greater connections, with those having few connections to other important people being at the bottom of the hierarchy \citep{bonacich1972norms}.

@@

\section{A Generalized Measure of Network Distinction}
Below, we present a measure of distinction centrality, determine the appropriate application of a scalar factor, and indicate the availability of an R program to facilitate easy implementation of the measure in further work.

@@

A formalization of this interaction between status and uniqueness allows for a measure of network distinction. Distinction as a theoretical concept is composed of status mediated by autonomy. A distinction measure can reflect this by considering an actor's status and then discounting their status by the status of their neighbors, independent of the actor. The more reliant an actor's neighbors are on the actor for their status, the lower the actor's ability to exert their influence, and thus the lower their distinction. In other words, autonomy is determined by how much one's neighbors rely on you to maintain their position in the network.

Distinction centrality can be measured using the following expression:
\begin{equation}
\beta_i=Sx_i-{\frac{1}{n_{j=i}}\sum_{{j}\in{A},i\neq{j}}x_{j}} 
\end{equation}

Let $\beta_i$ be the distinction centrality of a given individual \textit{i}. The first term captures the status of a given actor. Let \textit{x} be the eigenvector centrality of the designated individual. Let S be a scalar that adjusts the social value placed on status relative to constraint, as discussed below. The second term captures the element of autonomy. Removing individual \textit{i} from the network, compute the eigenvectors of the remaining individuals and sum those that are connected to individual \textit{i}, taking the average eigenvector of neighbors \textit{j} as the constraint confronting node \textit{i} in the network. This is substantively the same mechanism as conducted in a jackknife resampling procedure, except that rather than forming a sampling distribution, we are measuring local constraints. Finally, subtract the status of the average neighbor connected to \textit{i} from \textit{i}’s scalar-adjusted eigenvector centrality in the whole network.

When a node is an isolate, we set its eigenvector centrality to $x_i =0$. In the case of a network with no connections (such as an empty graph), some routines calculate it as $x_i =1.0$. However, given that it is still a network with no connections, setting isolates at $x_i =0$ is theoretically appropriate, considering that status is a reflection of connections. It would be inappropriate for an isolate to have the same status as the highest-status node in a large network. Additionally, given the second step of subtraction, it is only theoretically meaningful if set as $x_i =0$ in the case of distinction centrality.

\subsection{Scalar Adjustment}
First, a scalar is required because the relative value of status and autonomy is not inherently fixed in a social system. We might imagine situations where one desires more status at the expense of autonomy if the rewards to status are higher, or values autonomy more highly if the rewards to status are low. There cannot be an assumed unadjusted equivalence between the resulting eigenvectors of the two terms for this theoretical level.

Additionally, due to the hierarchical nature of networks, the average constraint in the network is greater than the average status. The high average constraint of their neighbors is often much larger than the impact of status. The result is that, at the level of the entire network, those who are deeply embedded among the most elite, without a basis of supporters, are very indistinct when considering the high constraints placed on them by their high-status alters. Thus, the calculated results without a scalar may fail to capture the appropriate value of status compared to the constraint.

While an arbitrary scalar is possible, it can yield erroneous results when applied to empirical networks. However, a scalar can be derived from the network itself due to the field positional interplay between status and constraint. Imagining two actors of equal status and personal constraint, one in an averagely more constrained system and the other in a less averagely constrained system, with average status otherwise equal. In such a case, the status of an actor in the more constrained system has a greater impact on their social distinction than in the less constrained system. With high constraint, status takes on greater importance in determining social distinction because it compels deference. In other words, it is the difference between a hierarchical system with and without constraints. Status with a weak constraint means a weaker compulsion of deference. The structure of constraint has a reverberative impact on social distinction beyond its contribution.

With this social understanding of how constraint interacts with the transmission of status, an endogenous scalar can be derived from the network. The average constraint of actors within the network divided by the average status of actors in the network creates a scalar that reflects the relationship between status and constraint.

\[S=\frac{\frac{1}{n_{j=i}}\sum_{{j}\in{A},i\neq{j}}x_{j}}{\overline{x_i}} \]

Under conditions of high average constraint and low average status, the scalar is large, indicating the significant role of status in determining social distinction. Under conditions where the average constraint is low and the average status is high, the scalar is weaker, reflecting a system in which status is less important in determining social distinction. Under very large-scale conditions, the averages of status and constraint should trend towards 1, weakening the necessity of the scalar. However, under constrained conditions, particularly in elite networks with missing data on the social scaffolding that created the network elites, the scalar is more important for explaining action because the range of eigenvector network measures is otherwise artificially expanded due to the missing data.

Without a scalar, networks with any degree of hierarchy see the average constraint overwhelm the average status because many low-status nodes are tied to high-status nodes. The advantage of this scalar is to ensure that the average scalar status in the network equals the average constraint in the network. Thus, total status is equal to total constraint, and from there it becomes a matter of distribution among members. From there, however, it enables comparability between networks that would otherwise be impossible. With the understanding that we have, through the scalar, adjusted for a network where one's combination of status and constraint is better or worse, networks can be compared so long as the unobserved social context within the network remains stable.

@@

\input{Tabs/star}
\input{Tabs/structfold.tex}
\input{Tabs/intraclust}
\input{Tabs/interclust}
\input{Tabs/circle}
\input{Tabs/kite}
\input{Tabs/clique.tex}
\input{Tabs/wheel}
\input{Tabs/circletriangle}
\input{Tabs/circlelong}

>>>>

The first aspect of social distinction is \textit{status}. Status refers to the internal stratification of all network members, along with the connections of deference and obligation. In the case of politics, status refers to the ability to access those who themselves have higher levels of access, with those most central to the access network having the authority to make decisions. This can be captured traditionally by measures of status such as eigenvector centrality \citep{bonacich1972factoring-03e}. Status is higher for actors when the people they are connected to are themselves influential, and accessibility is merely a political reframe of the context. The internal stratification of individuals is determined by who has greater connections to those with even greater connections, with those having few connections to other important people being at the bottom of the hierarchy \citep{bonacich1972factoring-03e}.

@@

\section{A Generalized Measure of Network Distinction}
In this section, we present our measure of distinction centrality, which incorporates a scalar adjustment factor to mitigate arbitrary scaling issues. We also indicate the availability of an \textit{R} custom function to facilitate the easy implementation of the measure in further work.

@@

Formalizing the interaction between status and uniqueness allows for a measure of network distinction. Distinction as a theoretical concept is composed of status mediated by autonomy. A distinction measure can reflect this by considering an actor's status and then discounting it by the counterfactual status of their neighbors when the focal actor is absent from the network. The more reliant an actor's neighbors are on the actor for their status, the lower the focal actor's ability to exert influence over them, and thus the lower their distinction. In other words, autonomy is determined by how much one's neighbors rely on you to maintain their advantageous position in the network.

Accordingly, we propose the following as a measure of distinction:

\begin{equation}
\beta_i= \alpha s_i - \kappa_i
\label{eq:distinction}
\end{equation}

Where $\beta_i$ is the distinction centrality score of actor \textit{i}, $s_i$ is actor $i$'s status, $\kappa_i$ is actor $i$'s network constraint (the obverse of autonomy), and $\alpha$ is a scalar adjustment factor to be discussed below. Each individual's status score $s_i$ is obtained from the vector $\mathbf{s}$ obtained from solving the standard linear eigenvector equation:

\begin{equation}
\lambda\mathbf{s}= \mathbf{A}\mathbf{s}
\label{eq:eigen}
\end{equation}

Where $\mathbf{A}$ is the network's adjacency matrix and $\mathbf{s}$ is the right dominant eigenvector of the matrix; status, therefore, is given by the usual eigenvector centrality score \citep{bonacich1972factoring-03e}, taken as the absolute value of each individual's score on the first eigenvector of the adjacency matrix. When a node is an isolate, either in the larger graph or in the node-deleted subgraph used to compute constraint scores, we set its status score to $s_i =0$.\footnote{In the case of a network with no connections (such as an empty graph), some routines calculate it as $s_i =1.0$. However, given that it is still a network with no connections, setting isolates at $s_i =0$ is theoretically appropriate, considering that status is a reflection of connections. It would be inappropriate for an isolate to have the same status as the highest-status node in a large network.}

The second term $\kappa_i$ in equation~\ref{eq:distinction} is a \textbf{localized induced centrality} score \citep{borgatti2020oxford-fa6}. It is given by the average status (eigenvector centrality) of $i$'s neighbors in the \textit{node-deleted subgraph} of the focal graph $G$ that excludes $i$:

\begin{equation}
    \kappa_i = \frac{\sum_{j \in N(i)}s_j^{G-i}}{|N(i)|}
\end{equation}

Where $|N(i)|$ is the cardinality of node $i$'s neighbor set (node $i$'s degree), and $s_j^{G-i}$ is the status of node $j$ in the graph that excludes node $i$.\footnote{The $\kappa$ term is meant to capture the element of autonomy. This is substantively the same mechanism as conducted in a jackknife resampling procedure, except that rather than forming a sampling distribution, we are measuring local constraints.} Thus the $\kappa_i$ component of equation~\ref{eq:distinction} is larger for nodes whose neighbors would gain a lot in status (on average) if they were to be removed from the network.

Understanding that both $s_i$ and $k_i$ are given in the (arbitrary) scaling produced by the eigenvector scoring procedure, it is desirable to re-scale the distinction centrality measure so that it has a natural zero point. As equation~\ref{eq:distinction} shows, we do this by multiplying each actor's status score by a scaling factor, which is given by:

\begin{equation}
    \alpha = \frac{\bar{\mathbf{\kappa}}}{\bar{\mathbf{s}}}
\end{equation}

Where $\bar{\mathbf{\kappa}}$ is the mean of the vector of $\kappa_i$ constraint scores for each actor in the network, and $\bar{\mathbf{\kappa}}$ is the mean of the vector of $s_i$ status scores for each actor in the network. As noted a desirable feature of rescaling the $s_i$ component of equation~\ref{eq:distinction} in this way is that in the case of a regular graph (e.g., a circle graph or a fully-connected clique), $\alpha s_i = \kappa_i$ and thus $\beta_i = 0$ for each node $i$ (see Tables~\ref{tab:tab:circle} and~\ref{tab:clique}).

@@

\input{Tabs/star}
\input{Tabs/structfold.tex}
\input{Tabs/intraclust}
\input{Tabs/interclust}
\input{Tabs/circle}
\input{Tabs/kite}
\input{Tabs/clique}
\input{Tabs/wheel}
\input{Tabs/circletriangle}
\input{Tabs/circlelong}
